---
title: "Stat 331 Final Project"
author: "Curtis LeQuang, Warren Woo, Johnny Young, Carter Coppes"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Data Description

We have chosen to analyze the relationship between a country's GDP per capita (in dollars) and a country's average daily income (in dollars). Information about each of these variables is contained in the two csv files `gdp_pcap.csv` and `mincpcap_cppp.csv`from the Gapminder data repository.

::: callout-note
**Definition:** GDP (or Gross Domestic Product) is a country's measure of the strength of their economy. GDP takes into account a country's consumption, investing, spending, and net exports over a certain period (usually one year) to calculate the total monetary value of goods and services in the country.

GDP per capita is just a standardized value for a country's GDP based on their population, which allows comparisons on economies dealing with different sizes of populations.
:::

Starting with the GDP data, each of the rows in the `gdp_pcap.csv` data frame represent a country in the world that tracks their GDP. The columns from 1800 to 2023 represent the years in which GDP per person data from the countries in the data set was collected. The columns from 2024 to 2100 are the *projected* GDP per person data from each country given the information we have collected in previous years.

The average income data is built very similarly to the GDP data, with each of the rows also representing a country that has accessible information on their average daily household incomes. The columns from 1800 to 2023 represent the years in which the average daily household income per capita was collected from the countries in the data set, and the columns from 2024 to 2100 represent the projected average daily household income per capita in the countries given the already collected information from past years.

It is important to note that in both of these data sets, the GDP per capita and average daily household income per capita are both measured in 2017 dollars, and both are inflation and price adjusted to standardize values between all of the countries (since economies in some countries are much larger than in others).

```{r setup}
#| code-fold: true
library(tidyverse)
library(kableExtra)

GDP <- read_csv(here::here("gdp_pcap.csv"))
income <- read_csv(here::here("mincpcap_cppp.csv"))
```

## Hypothesis

We are hypothesizing that a country's GDP per capita and its average daily household income per capita have a positive linear relationship. This would mean that as GDP per capita is increased, the average daily household income per capita will be predicted to also increase.

## Data Cleaning

As for the data cleaning process, these two data files started out messy. The column variables all had and **X** in front of each of the years, and the data entries were all classified as characters since values over 10,000 2017 dollars were notated with a **k** instead of as a numeric variable. So we first had to drop the **X** in each of the column variable names and change the **k** in each of the data entries to become the numeric notation.

Next, since we want our data to be in long format (where each country and each year has a value for GDP per capita and average daily household income per capita), we pivot both of these data sets to be longer in this format.

Also, the data in this set has been projected all the way to 2100, so we will only be looking at the data that has been physically collected already (from 1800 to 2023) so that we can run our own analysis using linear regression.

```{r data cleaning pt1}
#| code-fold: true
GDP_cleaned <- GDP |>
  select(country:'2023') |>
  mutate(across('1800':'2023', ~ as.numeric(str_replace_all(.x, "k", "e3")))) |>
  pivot_longer(cols = '1800':'2023',
               names_to = "year",
               values_to = "GDP per capita")
```

```{r data cleaning pt2}
#| code-fold: true
income_cleaned <- income |>
  select(country:'2023') |>
  mutate(across('1800':'2023', ~ as.numeric(str_replace_all(.x, "k", "e3")))) |>
  pivot_longer(cols = '1800':'2023',
               names_to = "year",
               values_to = "Average Daily Income per capita")
```

Finally, since both the GDP data set and the average daily household income data set contain the same countries, we just have to do an inner join between the two using the country as the key, and our resulting data set will have a country and year combination as each row, with a returned GDP per capita and an average daily household income per capita value. This will set us up to be able to do linear regression on our data.

```{r data joining}
#| code-fold: true
project_data <- GDP_cleaned |>
  inner_join(income_cleaned,
            by = join_by("country" == "country",
                         "year" == "year"
                         )
            )

head(project_data) |>
  kable()
```
